---
title: "HW3"
author: "Vincent Cholewa"
date: "10/17/2020"
output:
  word_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

```{r, warning=FALSE, message=FALSE}
library(TSA)
library(data.table)
library(vars)

hw3_data = read.csv('Currency Conversion Data.csv')
train = hw3_data[0:(length(hw3_data[,1])-8),]
train.ts = ts(train[,c(2,3,4,5)],start=c(2014,1),freq=52)
test = hw3_data[(length(hw3_data[,1])-7):length(hw3_data[,1]),]

#USD/EUR
EU.train = ts(train$USD.EU,start=c(2014,1),freq=52)
EU.test = test$USD.EU

#USD/GBP
GBP.train = ts(train$USD.GBP,start=c(2014,1),freq=52)
GBP.test = test$USD.GBP

#USD/AU
AU.train = ts(train$USD.AU,start=c(2014,1),freq=52)
AU.test = test$USD.AU

#USD/NZ
NZ.train = ts(train$USD.NZ,start=c(2014,1),freq=52)
NZ.test = test$USD.NZ
```

Question 1: Data Exploration (8 pts)

We already identified in past assignments that differenced currency conversion data tends to more closely resemble a stationary process. Knowing this, skip analyzing the original series, and for each series take the 1st order difference of the data.

Use relevant plots to evaluate assumptions of stationarity.

```{r}
par(mfrow=c(1,1))
plot(diff(EU.train),xlab="Years",ylab="",main="1st Ord Diff: USD/EUR",type="l")
plot(diff(GBP.train),xlab="Years",ylab="",main="1st Ord Diff: USD/GBP",type="l")
plot(diff(AU.train),xlab="Years",ylab="",main="1st Ord Diff: USD/AUD",type="l")
plot(diff(NZ.train),xlab="Years",ylab="",main="1st Ord Diff: USD/NZD",type="l")

par(mfrow=c(1,1))
acf(diff(EU.train), main="ACF D=1 USD/EUR", lag.max = 52)
acf(diff(GBP.train), main="ACF D=1 USD/GBP", lag.max = 52)
acf(diff(AU.train), main="ACF D=1 USD/AUD", lag.max = 52)
acf(diff(NZ.train), main="ACF D=1 USD/NZD", lag.max = 52)
```
Question 2: Univariate Analysis

2a. 
Use EACF plots (max order = 7) on the differenced data and iteration by AIC (original data, max order = 5, max differencing = 1, significance threshold of 2) to identify the best order for each series. How do the orders selected through each method compare?

```{r}
#eacf method
par(mfrow=c(2,1))
eacf(diff(EU.train), ar.max = 7, ma.max = 7)
eacf(diff(GBP.train), ar.max = 7, ma.max = 7)
eacf(diff(AU.train), ar.max = 7, ma.max = 7)
eacf(diff(NZ.train), ar.max = 7, ma.max = 7)

#iteration by aic method (max order = 5, diff = 1)
```


EU/USD
```{r message=FALSE, warning=FALSE}
eu_usd = function(p,d,q){
        mod = arima(EU.train, order=c(p,d,q), method="ML")
        current.aic = AIC(mod)
        df = data.frame(p,d,q,current.aic)
        names(df) <- c("p","d","q","AIC")
        
        #print(paste(p,d,q,current.aic,sep=" "))
        return(df)
}

eu_orders = data.frame(Inf,Inf,Inf,Inf)
names(eu_orders) = c("p","d","q","AIC")


for (p in 0:5){
  for (d in 0:1){
    for (q in 0:5) {
      possibleError <- tryCatch(
        eu_orders<-rbind(eu_orders,eu_usd(p,d,q)),
        error=function(e) e
      )
      if(inherits(possibleError, "error")) next

    }
  }
}
eu_orders = eu_orders[order(-eu_orders$AIC),]
tail(eu_orders,10) 
```

GBP/USD
```{r message=FALSE, warning=FALSE}
gbp_usd = function(p,d,q){
        mod = arima(GBP.train, order=c(p,d,q), method="ML")
        current.aic = AIC(mod)
        df = data.frame(p,d,q,current.aic)
        names(df) <- c("p","d","q","AIC")
        
        #print(paste(p,d,q,current.aic,sep=" "))
        return(df)
}

gbp_orders = data.frame(Inf,Inf,Inf,Inf)
names(gbp_orders) = c("p","d","q","AIC")


for (p in 0:5){
  for (d in 0:1){
    for (q in 0:5) {
      possibleError <- tryCatch(
        gbp_orders<-rbind(gbp_orders,gbp_usd(p,d,q)),
        error=function(e) e
      )
      if(inherits(possibleError, "error")) next

    }
  }
}
gbp_orders = gbp_orders[order(-gbp_orders$AIC),]
tail(gbp_orders,10) 
```

AUD/USD
```{r message=FALSE, warning=FALSE}
aud_usd = function(p,d,q){
        mod = arima(AU.train, order=c(p,d,q), method="ML")
        current.aic = AIC(mod)
        df = data.frame(p,d,q,current.aic)
        names(df) <- c("p","d","q","AIC")
        
        #print(paste(p,d,q,current.aic,sep=" "))
        return(df)
}

aud_orders = data.frame(Inf,Inf,Inf,Inf)
names(aud_orders) = c("p","d","q","AIC")


for (p in 0:5){
  for (d in 0:1){
    for (q in 0:5) {
      possibleError <- tryCatch(
        aud_orders<-rbind(aud_orders,aud_usd(p,d,q)),
        error=function(e) e
      )
      if(inherits(possibleError, "error")) next

    }
  }
}
aud_orders = aud_orders[order(-aud_orders$AIC),]
tail(aud_orders,10) 
```

NZD/USD
```{r message=FALSE, warning=FALSE}
nzd_usd = function(p,d,q){
        mod = arima(NZ.train, order=c(p,d,q), method="ML")
        current.aic = AIC(mod)
        df = data.frame(p,d,q,current.aic)
        names(df) <- c("p","d","q","AIC")
        
        #print(paste(p,d,q,current.aic,sep=" "))
        return(df)
}

nzd_orders = data.frame(Inf,Inf,Inf,Inf)
names(nzd_orders) = c("p","d","q","AIC")


for (p in 0:5){
  for (d in 0:1){
    for (q in 0:5) {
      possibleError <- tryCatch(
        nzd_orders<-rbind(nzd_orders,nzd_usd(p,d,q)),
        error=function(e) e
      )
      if(inherits(possibleError, "error")) next

    }
  }
}
nzd_orders = nzd_orders[order(-nzd_orders$AIC),]
tail(nzd_orders,10) 
```

b. (12 pts)
Regardless of what you observed above, fit the following ARIMA orders: EUR (2,1,3), GBP (1,1,1), AUD (1,0,3), NZD (2,0,2). Using these models, predict ahead 8 points. Plot these points with their 95% confidence intervals against the test data for each series.
```{r}
eu_arima = arima(EU.train, order = c(2,1,3), method = 'ML')
gbp_arima = arima(GBP.train, order = c(1,1,1), method = 'ML')
aud_arima = arima(AU.train, order = c(1,0,3), method = 'ML')
nzd_arima = arima(NZ.train, order = c(2,0,2), method = 'ML')

eu_pred = predict(eu_arima,   n.ahead=8)
gbp_pred = predict(gbp_arima, n.ahead=8)
aud_pred = predict(aud_arima, n.ahead=8)
nzd_pred = predict(nzd_arima, n.ahead=8)
```

```{r}
#You'll need this for plotting predictions
eu_whole = ts(hw3_data$USD.EU,start=c(2014,1),freq=52)
gbp_whole = ts(hw3_data$USD.GBP,start=c(2014,1),freq=52)
aud_whole = ts(hw3_data$USD.AUD,start=c(2014,1),freq=52) 
nzd_whole = ts(hw3_data$USD.NZ,start=c(2014,1),freq=52) 
times = time(eu_whole)
times.test = tail(times,8)
```

EU Plot
```{r}
eu_ub = eu_pred$pred + (1.96*eu_pred$se)
eu_lb = eu_pred$pred - (1.96*eu_pred$se)
eu_ymax = max(eu_ub)
eu_ymin = min(eu_lb)

n_eu = length(eu_whole)
nfit_eu = n_eu-8

par(mfrow=c(1,1))
plot(times[nfit_eu:n_eu], eu_whole[nfit_eu:n_eu], type='l', ylim=c(eu_ymin,eu_ymax), 
     main='EU/USD', 
     xlab="Time", 
     ylab="FX Rate")
points(times.test, eu_pred$pred, col='red')
lines(eu_ub,lty=3,lwd= 2, col="blue")
lines(eu_lb,lty=3,lwd= 2, col="blue")
```

GBP
```{r}
gbp_ub = gbp_pred$pred + (1.96*gbp_pred$se)
gbp_lb = gbp_pred$pred - (1.96*gbp_pred$se)
gbp_ymax = max(gbp_ub)
gbp_ymin = min(gbp_lb)

n_gbp = length(gbp_whole)
nfit_gbp = n_gbp-8

par(mfrow=c(1,1))
plot(times[nfit_gbp:n_gbp], gbp_whole[nfit_gbp:n_gbp], type='l', ylim=c(gbp_ymin,gbp_ymax), 
     main='GBP/USD', 
     xlab="Time", 
     ylab="FX Rate")
points(times.test, gbp_pred$pred, col='red')
lines(gbp_ub,lty=3,lwd= 2, col="blue")
lines(gbp_lb,lty=3,lwd= 2, col="blue")
```

AUD
```{r}
aud_ub = aud_pred$pred + (1.96*aud_pred$se)
aud_lb = aud_pred$pred - (1.96*aud_pred$se)
aud_ymax = max(aud_ub)
aud_ymin = min(aud_lb)

n_aud = length(aud_whole)
nfit_aud = n_aud-8

par(mfrow=c(1,1))
plot(times[nfit_aud:n_aud], aud_whole[nfit_aud:n_aud], type='l', ylim=c(aud_ymin,aud_ymax), 
     main='AUD/USD', 
     xlab="Time", 
     ylab="FX Rate")
points(times.test, aud_pred$pred, col='red')
lines(aud_ub,lty=3,lwd= 2, col="blue")
lines(aud_lb,lty=3,lwd= 2, col="blue")

```

NZD
```{r}
nzd_ub = nzd_pred$pred + (1.96*nzd_pred$se)
nzd_lb = nzd_pred$pred - (1.96*nzd_pred$se)
nzd_ymax = max(nzd_ub)
nzd_ymin = min(nzd_lb)


n_nzd = length(nzd_whole)
nfit_nzd = n_nzd-8

par(mfrow=c(1,1))
plot(times[nfit_nzd:n_nzd], nzd_whole[nfit_nzd:n_nzd], type='l', ylim=c(nzd_ymin,nzd_ymax), 
     main='NZD/USD', 
     xlab="Time", 
     ylab="FX Rate")
points(times.test, nzd_pred$pred, col='red')
lines(nzd_ub,lty=3,lwd= 2, col="blue")
lines(nzd_lb,lty=3,lwd= 2, col="blue")
```

c. (4 pts)

For each set of predictions, compute mean absolute percentage error (MAPE). How do they compare in accuracy?
```{r}
#EU/USD
eu_mape = 100*mean(abs(eu_pred$pred-EU.test)/EU.test)
eu_mape #0.5789397
#GBP/USD
gbp_mape = 100*mean(abs(gbp_pred$pred-GBP.test)/GBP.test)
gbp_mape #0.9954838       
#AUD/USD
aud_mape =100*mean(abs(aud_pred$pred-AU.test)/AU.test)
aud_mape #0.5598311
#NZD/USD
nzd_mape =100*mean(abs(nzd_pred$pred-NZ.test)/NZ.test)
nzd_mape #0.8846701
```

Question 3: Multivariate Analysis (28 pts)
a. (4 pts)
Identify the best VAR model that fits the four time series simultaneously according to AIC.
```{r}
par(mfrow=c(1,1))
fx.train = ts.union(EU.train, GBP.train, AU.train, NZ.train)

plot(fx.train, type='l')
acf(fx.train) #? why does this not plot 
pacf(fx.train)


VARselect(fx.train, lag.max = 20)$selection
plot(VARselect(fx.train, lag.max = 20)$criteria[1,], xlab='Order', ylab='AIC')
#p=2

mod_fx_1 = VAR(fx.train,lag.max=20,ic="AIC", type="none")
mod_fx_2 = VAR(fx.train,lag.max=20,ic="AIC", type="const")
mod_fx_3 = VAR(fx.train,lag.max=20,ic="AIC", type="trend")
mod_fx_4 = VAR(fx.train,lag.max=20,ic="AIC", type="both")

mod_fx_1$p #P = 2 
mod_fx_2$p #P = 2
mod_fx_3$p #P = 2
mod_fx_4$p #P = 2

mod_fx_hq = VAR(fx.train,lag.max=20,ic="HQ")
mod_fx_sc = VAR(fx.train,lag.max=20,ic="SC")
mod_fx_fpe = VAR(fx.train,lag.max=20,ic="FPE")

mod_fx_hq$p #2
mod_fx_sc$p #1 (BIC)
mod_fx_fpe$p #2
```

b. (12 pts)
Fit a VAR(2) model and check the model fitting using the multivariate ARCH test, the Jarque-Bera test and the Portmanteau test. State which assumptions are satisfied and which are violated
```{r}
#fit var(2) model
fx_model_var = VAR(fx.train, lag.max=20, p=2)
summary(fx_model_var)
fx_model_var_restrict = restrict(fx_model_var)
summary(fx_model_var_restrict)

#arch test
arch.test(fx_model_var)
#jarque-bera test
normality.test(fx_model_var)
#portmanteau test
serial.test(fx_model_var)
serialtest = serial.test(fx_model_var)
plot(serialtest)
```

c. (12 pts) 
Give the forecasts for the next 8 weeks of the series using a VAR(2) model. You don't need to plot them (but can if you'd like). Include confidence intervals.
```{r}
var2_pred = predict(fx_model_var, n.ahead=8)
var2_eu_pred = var2_pred[[1]]$EU.train[,c(1,4)]
var2_gbp_pred = var2_pred[[1]]$GBP.train[,c(1,4)]
var2_aud_pred = var2_pred[[1]]$AU.train[,c(1,4)]
var2_nzd_pred = var2_pred[[1]]$NZ.train[,c(1,4)]

#all results
var2_pred
#eu predictions & confident intervals
var2_eu_pred
#gbp predictions & confident intervals
var2_gbp_pred
#aud predictions & confident intervals
var2_aud_pred
#nzd predictions & confident intervals
var2_nzd_pred
```

Using mean absolute percentage error, compare the predictions for the four time series derived from the univariate analysis and multivariate analysis.
```{r}
eu_mape_var = 100*mean(abs(var2_eu_pred[,1]-EU.test)/EU.test)
gbp_mape_var = 100*mean(abs(var2_gbp_pred[,1]-GBP.test)/GBP.test)
aud_mape_var =100*mean(abs(var2_aud_pred[,1]-AU.test)/AU.test)
nzd_mape_var =100*mean(abs(var2_nzd_pred[,1]-NZ.test)/NZ.test)

eu_mape_var  #0.6303623
eu_mape  #0.5789397
gbp_mape_var #0.8045432
gbp_mape #0.9954838
aud_mape_var #1.264681
aud_mape #0.5598311
nzd_mape_var #0.6890983
nzd_mape #0.8846701
```


