---
title: "HW4"
author: "Vincent Cholewa"
date: "11/6/2020"
output: word_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message=FALSE}
library(quantmod)
library(tseries)
library(fGarch)
library(mgcv)
library(TSA)
library(rugarch)


tsla_data = read.csv('TSLA.csv')
#weekly returns 1/1/2012 - 4/28/2019
#View(tsla_data)
tsla_df = log(tsla_data[,2])
#length(tsla_df) 383
tsla.ts = ts(tsla_df,start=c(2012,1),freq=52)
data.growth = diff(tsla.ts)
```

Exploratory analysis (not graded):
```{r}
#Exploratory analysis
	#Plot original exchange rates
		plot(tsla.ts,type='l',main='TSLA Stock Price',ylab="Price")
		
	#Plot differenced series
		plot(data.growth,type='l',main='TSLA Stock Price',ylab="Diff Price")
		hist(data.growth)

	#ACF & PACF plots on original series
		par(mfcol=c(2,1))
		acf(tsla.ts,lag.max=52*3,main='ACF TSLA TS')
		pacf(tsla.ts,lag.max=52*3,main='PACF TSLA TS')

	#ACF & PACF plots on differenced series
		par(mfcol=c(2,1))
		acf(data.growth,ag.max=52*3,main='ACF Differenced TSLA TS')
		pacf(data.growth,ag.max=52*3,main='PACF Difference TSLA TS')
```

Question 1: ARIMA(p,d,q) (15 Points)
-With the growth rate data (first difference), use the iteration with the AIC metric selecting the minimum AIC (don't select a simpler model with higher AIC), select the best order of ARIMA model (max order 4,1,4).
-Plot residual and square residual ACFs and interpret.
-Perform and interpret tests for serial correlation and heteroscedasticity on model residuals. 

```{r}
## Fit ARIMA
tsla_model = function(p,d,q){
    mod = arima(data.growth, order=c(p,d,q), method="ML")
    current.aic = AIC(mod)
    df = data.frame(p,d,q,current.aic)
    names(df) <- c("p","d","q","AIC")
    print(paste(p,d,q,current.aic,sep=" "))
    return(df)
}

orders = data.frame(Inf,Inf,Inf,Inf)
names(orders) <- c("p","d","q","AIC")

for (p in 0:4){
    for(d in 0:1) {
        for (q in 0:4) {
            possibleError <- tryCatch(
                orders<-rbind(orders,tsla_model(p,d,q)),
                error=function(e) e
            )
            if(inherits(possibleError, "error")) next

        }
    }
}
orders <- orders[order(-orders$AIC),]
min(orders);tail(orders)
#arima(4,0,2) aic=-1003.803

tsla_arima = arima(data.growth, order=c(4,0,2))
tsla_arima      


## Residual Analysis
tsla_resids = resid(tsla_arima)[-1]
#plot of residuals
	par(mfcol=c(2,1))
	plot(tsla_resids,main='Residuals of TSLA ARIMA Fit')
	plot(tsla_resids^2,main='Squared Residuals of TSLA ARIMA Fit')

#acf plots
acf(tsla_resids, main="Residuals of ARIMA Fit")
acf(tsla_resids^2 ,main="Squared Residuals of ARIMA Fit")

# for serial correlation
Box.test(tsla_resids,lag=7,type='Ljung',fitdf=6) #df = arima(4,0,2), lag = df + 1
#p-value = 0.3182; fail to reject null = plausibly uncorrelated residuals

# for arch effect
Box.test((tsla_resids)^2,lag=7,type='Ljung',fitdf=6) #df = arima(4,0,2), lag = df + 1
#p-value = 0.01551; reject the null = correlated residuals 
arch.test(tsla_arima)
```

Question 2: ARMA(p,q)GARCH(m,n) (20 Points)
-With the growth rate data (first difference) and using the multi-step refinement method (see piazza post @628) with minimum BIC starting with ARMA order calculated in Q1, find the best ARMA(P,Q)-GARCH(M,N) order pair. 

The order of the process is:
1. Select an ARMA order (same as we've done so far)
2. With the ARMA order selected in 1, select a GARCH order
3. With the new GARCH order, select the best ARMA order
4. Of the potentially 2 ARMA-GARCH orders now found, choose the one with the lowest BIC.

```{r}
#Initial GARCH Order
#ARIMA-GARCH GARCH order
test_modelAGG <- function(m,n){
  spec = ugarchspec(variance.model=list(garchOrder=c(m,n)),
                    mean.model=list(armaOrder=c(4,2), 
                                    include.mean=T), distribution.model="std")    
  fit = ugarchfit(spec, data.growth, solver = 'hybrid')
  current.bic = infocriteria(fit)[2]
  df = data.frame(m,n,current.bic)
  names(df) <- c("m","n","BIC")
  print(paste(m,n,current.bic,sep=" "))
  return(df)
}

garch_orders = data.frame(Inf,Inf,Inf)
names(garch_orders) <- c("m","n","BIC")


for (m in 0:2){
  for (n in 0:2){
    possibleError <- tryCatch(
      garch_orders<-rbind(garch_orders,test_modelAGG(m,n)),
      error=function(e) e
    )
    if(inherits(possibleError, "error")) next
  }
}
garch_orders <- garch_orders[order(-garch_orders$BIC),]
tail(garch_orders); min(garch_orders) #arch(1,2) with bic = -2.585301

#ARMA update
#ARIMA-GARCH ARIMA order
test_modelAGA <- function(p,q){
  spec = ugarchspec(variance.model=list(garchOrder=c(1,2)),
                    mean.model=list(armaOrder=c(p,q), 
                                    include.mean=T), distribution.model="std")    
  fit = ugarchfit(spec, data.growth, solver = 'hybrid')
  current.bic = infocriteria(fit)[2]
  df = data.frame(p,q,current.bic)
  names(df) <- c("p","q","BIC")
  print(paste(p,q,current.bic,sep=" "))
  return(df)
}

arma_orders = data.frame(Inf,Inf,Inf)
names(arma_orders) <- c("p","q","BIC")


for (p in 0:4){
  for (q in 0:4){
    possibleError <- tryCatch(
      arma_orders<-rbind(arma_orders,test_modelAGA(p,q)),
      error=function(e) e
    )
    if(inherits(possibleError, "error")) next
  }
}
arma_orders <- arma_orders[order(-arma_orders$BIC),]
tail(arma_orders); min(arma_orders) #arma(4,4); bic = -2.634431


#GARCH update
test_modelAGG <- function(m,n){
  spec = ugarchspec(variance.model=list(garchOrder=c(m,n)),
                    mean.model=list(armaOrder=c(4,4), 
                                    include.mean=T), distribution.model="std")    
  fit = ugarchfit(spec, data.growth, solver = 'hybrid')
  current.bic = infocriteria(fit)[2]
  df = data.frame(m,n,current.bic)
  names(df) <- c("m","n","BIC")
  print(paste(m,n,current.bic,sep=" "))
  return(df)
}

orders_arma_arch = data.frame(Inf,Inf,Inf)
names(orders_arma_arch) <- c("m","n","BIC")


for (m in 0:2){
  for (n in 0:2){
    possibleError <- tryCatch(
      orders_arma_arch<-rbind(orders_arma_arch,test_modelAGG(m,n)),
      error=function(e) e
    )
    if(inherits(possibleError, "error")) next
  }
}
orders_arma_arch <- orders_arma_arch[order(-orders_arma_arch$BIC),]
tail(orders_arma_arch); min(orders_arma_arch) #arch(1,1) bic = -2.644133
#best model is arma(4,4)garch(1,1)
```
Fit this model.
-Fully write out the equation. You can simply state the level of differencing Y represents in the data for simplicity sake.
-Perform goodness of fit tests on this model and interpret. (Hint: summary of a garchFit model performs them all, you just need to interpret).
```{r}
#fit and write out model using garchFit
tsla_archFit_ts = garchFit(~ arma(4,4)+ garch(1,1), data=data.growth, trace = FALSE)
summary(tsla_archFit_ts) #goodness of fit also shown in output

#Forumula
##Differenced data
##yt = 0.00426141 - 0.75959443yt-1 - 0.19646565yt-2 + 0.57980867yt-3 + 0.09368774yt-4 + zt + 0.76189675zt-1 + 0.20717926zt-2 - 0.54108576zt-3 + 0.01292682zt-4 
##zt = sigmatRt
##sigmat^2 = 0.00026663 + 0.06429848zt^2-1 + 0.8702856sigmat^2-1
```

Question 3: Forecasting (15 Points)
Using the rolling forecasting method (for each pred, fit A-G model to all points prior), forecast last 40 points.
-Calculate MAPE, and Precision
```{r}
tsla_spec_1 = ugarchspec(variance.model=list(garchOrder=c(1,1)),
                 mean.model=list(armaOrder=c(4, 4),
                 include.mean=T), distribution.model="std")

n = length(data.growth)
data.test = data.growth[(length(data.growth)-39):length(data.growth)]
data.train = data.growth[1:(length(data.growth)-40)]


#Prediction of the return time series and the volatility sigma
nfore = length(data.test)
fore.series.1 = NULL
fore.sigma.1 = NULL

for(f in 1:nfore)
        {
        #Fit models
        data = data.train
        if(f>2)
                data = c(data.train,data.test[1:(f-1)])  
        final.model = ugarchfit(tsla_spec_1, data, solver = 'hybrid')   
                #Forecast
        fore = ugarchforecast(final.model, n.ahead=1)
        fore.series.1 = c(fore.series.1, fore@forecast$seriesFor)
        fore.sigma.1 = c(fore.sigma.1, fore@forecast$sigmaFor)
}
length(fore.series.1)
###Compute Accuracy Measures 
## Mean Absolute Percentage Error (MAPE)
mean(abs(fore.series.1 - data.test)/abs(data.test))
##Precision Measure (PM)
sum((fore.series.1 - data.test)^2)/sum((data.test-mean(data.test))^2)
```

-Overlay the draw an overlay of the predicted points on the original series in that range.
```{r}
##plot points
ymin = min(c(as.vector(data.test), fore.series.1), na.rm =T)
ymax = max(c(as.vector(data.test), fore.series.1), na.rm =T)
data.plot = data.test
names(data.plot)="Fore"
n = length(data.growth)
plot(ts(data.growth[c(n-39):n]),type="l", ylim=c(ymin,ymax), xlab="Time (Weekly)", ylab="Diff: Return Price")
points(data.test, lwd=2, col='blue')
points(fore.series.1, lwd=2, col='red')

length(fore.series.1)

##plot variance
plot(data.growth[c(n-40):n]^2,type="l", ylim=c(ymin,ymax), xlab="Time", ylab="Return Price")
data.plot$Fore=fore.sigma.1^2
points(data.frame(data.plot), lwd=2, col='blue')
```